---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: true
---

author : Gerrit Versteeg

date   : Oct 16th, 2016


## Loading and preprocessing the data

Because the dataset (in ZIP-format) is already in the contents of the forked
repo, we do not have to download the file again. _(Note: the original data was
downloaded in RPeng's Repo on Feb 11th, 2014.)_  

After loading the dplyr package silently, the first step to take is unzipping
'activity.zip' into a ./data directory and then read the resulting 
'activity.csv' into a dataset called 'DF-raw' using dplyr. A peek of the content
is shown.

```{r loading,ECHO = TRUE}
library("dplyr", warn.conflicts=FALSE)             ## load dplyr silently
if (file.exists("./data/activity.csv")) {          ## check if file exists
        unlink("./data/activity.csv")              ## if so -> delete 
}
unzip("./activity.zip", exdir = "./data")          ## unzip into data dir
DF_raw <- tbl_df(read.csv("./data/activity.csv"))  ## load into dataset/tibble
DF_raw                                             ## and take a peek
```

Next step is to pre-process the data for analysis. Obviously the date is 
formatted as factor, inhibiting processing in weekdays and weekend later on.
So we need to reformat that variable to the 'Date' format. 

```{r Preprocessing,ECHO=TRUE}
DF_raw$date <- as.Date(strptime(DF_raw$date, "%Y-%m-%d"))  ## reformat to date
class(DF_raw$date)                                         ## show class
```


## What is the mean number of daily steps taken?

First we calculate the total number of steps taken each day. Using dplyr we
group the dataset by date and sum the steps into a column named 'total'.

```{r totalstepsdaily, ECHO=TRUE}
DF_tsd <-                                    ## dplyr-code to create DF-tsd
        DF_raw %>%                           ## use DF-raw to
        select(steps, date) %>%              ## select relevant columns
        group_by(date) %>%                   ## group them by date
        summarize(total = sum(steps)) %>%    ## calculate the sum of steps
        print()                              ## and take a peek
```

_Note: you can see that there are days with no observations at all
(like Oct 1st, 2012 and Oct 8th, 2012)._


Now let's plot a histogram of the frequency of the number of steps taken daily 
to get an idea of the most frequent values.

```{r tsd-hist, ECHO=TRUE}
with(DF_tsd, hist(total,                              ## freq. daily steps
     main = "Histogram - Number of daily steps",      ## set main title
     xlab = "Number of daily steps",                  ## set label X-axis
     ylab = "Frequency"))                             ## set label Y-axis
```

Apparently the most frequent number of daily steps lies between **10k and 15k** 
steps per day.

Finally we need to take a look at the mean and the median of the 
daily total of steps. Using summary will automatically separate the NA's and
show us some more info.

```{r tsd-mean, ECHO=TRUE}
sum_raw <- as.integer(summary(DF_tsd$total))  ## store for later comparison
summary(DF_tsd$total)                         ## show summary with mean & median
```

The mean of all daily steps is **10.770** steps per day, while the median is
**10.760** steps/day. Also visible is the fact that there are 8 days with no 
observations (registration of steps) at all.


## What is the average daily activity pattern?
Now we want to take a look at the daily activity pattern using the average
number of steps taken in each time interval. Looking at the variable 'interval'
we see that the intervals vary from 0 (00:00-04:59) up to 2355 (23:55-23:59).
These 288 (60 minutes * 24 hours / 5 minutes) values will be the time-slots
used for the activity pattern, while we compute the mean of the daily number
of steps for each of these intervals (ignoring any NA-values). 

```{r dailyactivitypattern, ECHO=TRUE}
DF_dap <-                                    ## dplyr-code to create DF-dap
        DF_raw %>%                           ## use DF-raw to
        select(steps, interval) %>%          ## select relevant columns
        group_by(interval) %>%               ## group them by interval
        summarize(avg = mean(steps,          ## calculate the mean of each
                  na.rm = TRUE)) %>%         ## interval across days (excl. NA's)
        print()                              ## and take a peek
```

The peek shows the mean steps taken for the first intervals.  As can be
expected, there is not much activity going on just after midnight.


Now for the activity pattern of an average day, let's plot this dataset as a
time-series, showing the average (across days) of steps taken during each 
specific interval. 

```{r dap-plot, ECHO=TRUE}
with(DF_dap, plot(interval, avg,             ## plot DF_dap (avg.steps˜interval)
     type ="l",                              ## using a line-plot
     main = "Daily Activity Pattern",        ## set main title
     xlab = "5-minute interval",             ## set label X-axis
     ylab = "Average number of steps"))      ## set label Y-axis
```

Looking at the resulting pattern, there seems to be a clear peak in one of the
5-minute intervals in the morning (over **200** steps). This might be due to
some kind of commuting to/from work, but the same pattern does not re-occur at
the end of the workday. So maybe it indicates a sort of 'work out' in the form
of a morning run or walk each day prior to going to work.

Because we like to see in which specific interval this peak lies, 
we select and print the row where 'average' equals the maximum value.

```{r dap-max, ECHO=TRUE}
filter(DF_dap, avg == max(avg))  ## select row of interval with max # of steps
```

The result shows that interval **835** (08:35-08:39) contains the highest number
of steps averaged across days. Namely roughly **206** steps. 


## Imputing missing values
Some of the measures are missing, in other words; during some intervals or
even days, the steps taken were not registered. These missing measurements are 
noted as NA's. Let's first look at the number of missing values (NA's).

```{r NA-count1, ECHO=TRUE}
sum(is.na(DF_raw$steps))        ## calculate NA's in Steps variable
```

Apparently there are **2304** missing values. Just to be sure that NA's only
occur in de steps variable, we will check the NA-count over all variables in
the dataset.

```{r NA-count2, ECHO=TRUE}
sum(is.na(DF_raw))              ## calculate NA's in complete dataframe
```

Clearly the number of NA's in the complete dataset is the same as the number
of NA's in de steps variable (i.e. 2304). Therefor NA's only occur in the
variable 'steps'.

####Imputing strategy
What will we replace the missing values with? Looking at the distribution 
with the peak around 8:30, replacement of NA's with the mean of a complete day
would 'ruin' the pattern and because in some cases complete days are missing
would result in eight additional NA's. A better solution seems to be to impute
values that coincide with the mean of that specific interval.


So, we will impute the missing values with the mean of the interval at hand. 
Using dplyr's mutate to replace the NA with 'mean(steps)' would result in the
mean of **all** steps in the dataframe, if we leave the dataset as is 
(i.e. ungrouped). To fill it with the means per interval, we need to group
the dataset using group_by(interval). Calling mean(steps) will now return the 
mean for the interval rather than the overall mean. After determining the mean, 
we remove the grouping because the resulting dataset does not need grouping per 
interval.

```{r Imputing NAs, ECHO=TRUE}
DF_ina <-                                     ## create new dataset
        DF_raw %>%                            ## derived from DF_raw
        group_by(interval) %>%                ## grouped to get mean/interval
        mutate(steps = ifelse(is.na(steps),   ## replace steps if value is NA
        as.integer(mean(steps, na.rm=TRUE)),  ## with the mean of the interval
        steps)) %>%                           ## or leave steps alone.
        group_by() %>%                        ## Remove the interval grouping
        print                                 ## and take a peek
```

We can see that the NA's are replaced by the interval means.


Next we recalculate the total number of steps taken each day. Using dplyr we
group the dataset by date and sum the steps into a column named 'total'.

```{r totalstepsdaily-nai, ECHO=TRUE}
DF_tsdina <-                                 ## dplyr-code to create DF-tsdina
        DF_ina %>%                           ## use DF-ina to
        select(steps, date) %>%              ## select relevant columns
        group_by(date) %>%                   ## group them by date
        summarize(total = sum(steps)) %>%    ## calculate the sum of steps
        print()                              ## and take a peek
```

Now let's regenerate the histogram to get an idea of the change in the frequency
of the number of steps taken daily.

```{r tsd-hist-ina, ECHO=TRUE}
with(DF_tsdina, hist(total,                           ## freq. daily steps
     main = "Histogram - Number of daily steps",      ## set main title
     xlab = "Number of daily steps",                  ## set label X-axis
     ylab = "Frequency"))                             ## set label Y-axis
```

Apparently the most frequent number of daily steps still lies between **10k and 
15k** steps per day, but the number of observations between 10k and 15k has risen
from approx. **28** to approx. **36**. Imputing NA's with the interval mean 
apparantly tightened the number of daily steps towards the mean, as can be 
expected.

Finally we need to take a another look at the mean and the median of the 
daily total of steps, again using summary.

```{r tsdina-mean, ECHO=TRUE}
sum_ina <- as.integer(summary(DF_tsdina$total)) ## store for comparison
summary(DF_tsdina$total)                        ## show summary for mean/median
```

The new mean is **10.750** steps/day, while the new median adds up to **10.640**
steps/day.

Comparing the results:

|   | Min | Median | Mean |  Max  | NAs |
|----|----|----|----|----|----|
| Original |`r sum_raw[[1]]` |`r sum_raw[[3]]` |`r sum_raw[[4]]` |`r sum_raw[[6]]` |`r sum_raw[[7]]` |
| Imputed  |`r sum_ina[[1]]` |`r sum_ina[[3]]` |`r sum_ina[[4]]` |`r sum_ina[[6]]` | |

After imputing the interval mean, obviously the mean of all daily steps appears 
to have changed from **10.770 to 10.750** steps per day, while the median changed
from **10.760 to 10.640** steps/day. Apparantly the imputed values are lower than
the original mean and therefore drag down the mean. No NA's remain after imputing.

## Are there differences in activity patterns between weekdays and weekends?

To be able to distinguish between weekdays and weekend days, we need to add
another variable 'ww' to the dataset DF_ina (note: this is the dataset with the 
imputed values). Using the weekdays function on the variable 'date'
(that was reformated into date whilst cleaning the data at the start) 
we will check to see whether the data is a 'saturday' or a 'sunday". If so,
we set 'ww' to 'weekend' otherwise to 'weekday'.


```{r create-ww-column, ECHO=TRUE}
WE <- c("Saturday", "Sunday")                ## make vector with 'weekend'-days
DF_inaw <-                                   ## dplyr-code to create DF-inaw
        DF_ina %>%                           ## use DF-ina to
        mutate(ww = as.factor(ifelse(        ## add a factor column 'ww'
                weekdays(date) %in% WE,      ## check if weekday = Sat or Sun
                           "weekend",        ## then fill it with "weekend"
                           "weekday"))) %>%  ## else fill it with "weekday"
        print                                ## and take a peek
```

We determine the new activity pattern (now with imputed NA's) we group the 
dataset DF_inaw on 'ww' and 'interval' because we now need to calculate the mean
number of daily steps for each type of day (weekday or weekend) **and** 
for each interval across all days.

```{r dailyactivitypattern-ina, ECHO=TRUE}
DF_dapina <-                                 ## dplyr-code to create DF-dapina
        DF_inaw %>%                          ## use DF-inaw to
        select(steps, interval, ww) %>%      ## select the relevant columns
        group_by(ww, interval) %>%           ## group them by ww & interval
        summarize(avg = mean(steps, 
                  na.rm = TRUE)) %>%         ## calculate the mean across days
        print                                ## and take a peek
```

The new dataset (DF_dapina) with the daily average steps per interval has 576
rows now, because we split the complete dataset into weekdays and weekends, 
resulting in a 288-interval set for weekdays and an 288-interval set for
weekends.
Now that we have the dataset that we need, we can start to plot this
as a time-series in two frames (one for weekend and one for weekdays), showing
the average (across days) of steps taken during each specific interval of these
two types of days. 

*Please note that I use ggplot2 for this plot, because I like the way it shows
multiple facets. This differs a bit from the example plot given by our mentor. 
He stated that is not needed to reproduce the figure exactly as is. As long as
the comparison is facilitated in the same manner. So I hope you don't mind*

```{r dapw-plot, ECHO=TRUE}
library(ggplot2)
g <- ggplot(DF_dapina, aes(interval, avg))           ## setup graphic object
g+geom_path()+                                       ## plot the path
        facet_grid(ww~.) +                           ## setup grid (2row x 1col)
        xlab("5-minute interval") +                  ## label X-axis
        ylab("Average number of steps") +            ## label Y-axis
        ggtitle("Daily Activity Patterns compared")  ## title the graph
```

Activity appears to start **earlier** in weekdays (around 05:00) than in 
weekends (around 07:30). The **'workout' pattern** is still visible during
weekdays, refering to the peak around 8:30 followed by lower activity during the
remainder of the day. While during weekends, there is **more** activity going on
that is **spread more evenly** across the day.

