---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: true
---

author : Gerrit Versteeg

date   : Oct 12th, 2016


## Loading and preprocessing the data

Because the dataset (in ZIP-format) is already in the contents of the forked
repo, we do not have to download the file again.  
_Note: the original data was downloaded in RPeng's Repo on Feb 11th, 2014._  

First step to take is unzipping 'activity.zip' into a ./data directory and then
read the resulting 'activity.csv' into a tibble called 'DF-raw' using dplyr. 
A peek of the content is shown.

```{r loading,ECHO = TRUE}
library("dplyr", warn.conflicts=FALSE)
if (file.exists("./data/activity.csv")) {
        unlink("./data/activity.csv")
}
unzip("./activity.zip", exdir = "./data")
DF_raw <- tbl_df(read.csv("./data/activity.csv"))
DF_raw
```

Next step is to pre-process the data for analysis. Obviously the date is 
formatted as factor, inhibiting processing in weekdays and weekend. So we need
to reformat that variable to the 'Date' format. 

```{r Preprocessing,ECHO=TRUE}
DF_raw$date <- as.Date(strptime(DF_raw$date, "%Y-%m-%d"))
class(DF_raw$date)
```


## What is mean total number of steps taken per day?

First we calculate the total number of steps taken each day. Using dplyr we
group the tibble by date and sum the steps into a column named 'total'.

```{r totalstepsdaily, ECHO=TRUE}
DF_tsd <-                                    ## dplyr-code to create DF-tsd
        DF_raw %>%                           ## use DF-raw to
        select(steps, date) %>%              ## select relevant columns
        group_by(date) %>%                   ## group them by date
        summarize(total = sum(steps)) %>%    ## calculate the sum of steps
        print()                              ## and take a peek
```

Now let's plot a histogram of the frequency of the number of steps taken daily 
to get an idea of the most frequent values.

```{r tsd-hist, ECHO=TRUE}
with(DF_tsd, hist(total,                              ## freq. daily steps
     main = "Histogram - Number of daily steps",      ## set main title
     xlab = "Number of daily steps",                  ## set label X-axis
     ylab = "Frequency"))                             ## set label Y-axis
```

Apparently the most frequent number of daily steps lies between 10k and 15k 
steps per day.

Finally we need to take a look at the mean and the median of the 
daily total of steps. Using summary will automatically separate the NA's and
show us some more info.

```{r tsd-mean, ECHO=TRUE}
sum_raw <- summary(DF_tsd$total)
sum_raw
```

Obviously the mean of all daily steps appears to be 10.770 steps per day,
while the median is 10.760 steps/day. Also visible is the fact that there are
8 days with no step registration at all.


## What is the average daily activity pattern?
Now we want to take a look at the daily activity pattern using the average
number of steps taken in each time interval. Looging at the variable 'interval'
we see that the intervals vary from 0 (00:00-04:59) up to 2355 (23:55-23:59).
These 288 (60*24/5) values will be the time-series used for the activity
pattern, while we compute the mean of the daily number of steps for each these
intervals (ignoring any NA-values). 

```{r dailyactivitypattern, ECHO=TRUE}
DF_dap <-                                    ## dplyr-code to create DF-dap
        DF_raw %>%                           ## use DF-raw to
        select(steps, interval) %>%          ## select relevant columns
        group_by(interval) %>%               ## group them by interval
        summarize(avg = mean(steps, 
                  na.rm = TRUE)) %>%         ## calculate the mean across days
        print()                              ## and take a peek
```

Now let's plot this as time-series, showing the average (across days) of 
steps taken during each specific interval of a day. 

```{r dap-plot, ECHO=TRUE}
with(DF_dap, plot(interval, avg,             ## plot DF_dap (avg.steps˜interval)
     type ="l",                              ## using a line-plot
     main = "Daily Activity Pattern",        ## set main title
     xlab = "5-minute interval",             ## set label X-axis
     ylab = "Average number of steps"))      ## set label Y-axis
```

Looking at the resulting pattern, there seems to be a clear peak in the morning,
that might indicate a morning run or walk each day prior to going to work. 
If we like to see in which specific interval the peak lies, we select and print 
the row where the average equals the maximum value.

```{r dap-max, ECHO=TRUE}
filter(DF_dap, avg == max(avg))
```

The result shows that interval 835 (08:35-08:39) contains the highest number of
steps averaged across days. Namely roughly 206 steps. 


## Imputing missing values
Some of the measures are missing, in other words; during some intervals or
even days, the steps taken were not registered. These missing measurements are 
noted as NA's. Let's first look at the number of missing values (NA's).

```{r NA-count1, ECHO=TRUE}
sum(is.na(DF_raw$steps))        ##calculate NA's in Steps variable
```

Apparently there are 2304 missing values. Just to be sure that we NA's only
occur in de steps variable, we will check hat the NA-count is over the complete
tibble.

```{r NA-count2, ECHO=TRUE}
sum(is.na(DF_raw))              ##calculate NA's in complete dataframe
```

Clearly the number of NA's in the complete tibble is the same as the number
of NA's in de steps variable (i.e. 2304).

####Imputing strategy
What will be the replacements for missing values?. Looking at the distribution with the peak around 8:30, replacement of NA's with the mean of a complete day would ruin the pattern and because in some cases complete days are missing would result in an additional NA. A better solution seems to be to impute values that coincide with the mean of that specific interval. This will retain the activity pattern across the day.

When imputing the missing values, we need to do that with the mean of the interval at hand. Using dplyr's mutate and replace the NA with 'mean(steps') would result in the mean of all steps in the dataframe. That is why we use group_by(interval). Calling mean(steps) will then return the mean for the interval rather than the overall mean.

```{r NA-imputing, ECHO=TRUE}
DF_nas <-                                     ## create new tibble
        DF_raw %>%                            ## derived from DF_raw
        group_by(interval) %>%                ## grouped to get mean/interval
        mutate(steps = ifelse(is.na(steps),   ## replace steps if value is NA
        as.integer(mean(steps, na.rm=TRUE)),  ## with the mean of the interval
        steps)) %>%                           ## or leave steps alone.
        group_by() %>%                        ## Remove the interval grouping
        print                                 ## and take a peek
```

We can see that the NA's are replaced by the interval means.
Next we recalculate the total number of steps taken each day. Using dplyr we
again group the tibble by date and sum the steps into a column named 'total'.

```{r totalstepsdailynas, ECHO=TRUE}
DF_tsdnas <-                                 ## dplyr-code to create DF-tsd
        DF_nas %>%                           ## use DF-nas to
        select(steps, date) %>%              ## select relevant columns
        group_by(date) %>%                   ## group them by date
        summarize(total = sum(steps)) %>%    ## calculate the sum of steps
        print()                              ## and take a peek
```

Now let's regenerate the histogram of the frequency of the number of steps taken
daily to get an idea of the change in the most frequent values.

```{r tsd-hist-nas, ECHO=TRUE}
with(DF_tsdnas, hist(total,                           ## freq. daily steps
     main = "Histogram - Number of daily steps",      ## set main title
     xlab = "Number of daily steps",                  ## set label X-axis
     ylab = "Frequency"))                             ## set label Y-axis
```

Apparently the most frequent number of daily steps still lies between 10k and 
15k steps per day, but the number of observations between 10k and 15k has risen
from approx. 28 to approx. 36. Using the interval mean clearly tightened the
number of daily steps towards the mean, as can be expected.

Finally we need to take a another look at the mean and the median of the 
daily total of steps, again using summary.

```{r tsdnas-mean, ECHO=TRUE, results="asis"}
library(xtable)
sum_nas <- summary(DF_tsdnas$total)
dt <- t(matrix(unname(sum_raw)))
dt <- rbind(dt, c(unname(sum_nas), "0"))
xt <- xtable(dt,
             caption="Comparison of summaries",
             digits=c(0,0,0,0,0,0,0,0),
             auto = TRUE)
colnames(xt) <- c(" ", " Min ", " 1st Qu ", " Median ", " Mean ", " 3rd Qu  ", "  NAs   ")
rownames(xt) <- c("Original", "Imputed")
align(xt) <- "p{2cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}"
print(xt, type = "html", html.table.attributes = "cellspacing='10'")
```

|  | Mean | Median |
|----|----|----|
| Original |`r as.integer(mean(DF_tsdnas$total))` | `r as.integer(median(DF_tsdnas$total))`|
| Imputed |`r as.integer(mean(DF_tsd$total))` | `r as.integer(median(DF_tsd$total))`|


After imputing the interval mean, obviously the mean of all daily steps appears 
to have changed from 10.770 to 10.750 steps per day, while the median changed
from 10.760 to 10.640 steps/day. Apparantly the imputed values are low and drag
down the mean. No NA's remain now.

## Are there differences in activity patterns between weekdays and weekends?
